---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading libraries
```{r libraries}
library(dplyr)
library(ggplot2)
library(lattice)
```


## Loading and preprocessing the data
#### Reading data
```{r readData, echo=TRUE}
if(!file.exists("activity.csv")) {
        unzip("activity.zip")
}

data <- read.csv("activity.csv", header=TRUE,
                 colClasses=c("integer", "Date", "integer"))
```

#### Defining some useful functions
```{r functions, echo=TRUE}
# converting intervals to decimal notation and vice versa
totime <- function(numChar) {
        num <- as.numeric(numChar)
        hour <- floor(num / 100)
        min_decimal <- (num - hour * 100) / 60
        hour + min_decimal
}
tochar <- function(decimalTime) {
        hour <- floor(decimalTime)
        minutes <- round((decimalTime - hour) * 60, digits=0)
        paste(as.character(hour), as.character(minutes), sep=":")
}
```

#### Processing the data
```{r processing, echo=TRUE}
data_day <- data %>%
                group_by(date) %>%
                summarise(dailyCount=sum(steps, na.rm=TRUE))
data_interval <- data %>%
                group_by(interval) %>%
                summarise(intervalCount=mean(steps, na.rm=TRUE))
data_interval$interval <- totime(data_interval$interval)
```


## What is mean total number of steps taken per day?

Here's a histogram of the data.
```{r histMeanMedian, echo=TRUE}
# calculating the mean and median
mean.day <- mean(data_day$dailyCount)
median.day <- median(data_day$dailyCount)

g <- ggplot(data_day, aes(x=dailyCount))
g +
        geom_histogram() +
        scale_x_log10() +
        geom_vline(xintercept=mean.day, size=2, color="red") +
        geom_vline(xintercept=median.day, size=2, color="blue") +
        xlab("Number of steps") +
        ylab("Frequency") +
        labs(title="Histogram of Steps Taken")
```

In the above plot, the mean of the number of steps taken is `r as.integer(mean.day)` shown in **red** and the median of the number of steps taken is `r as.integer(median.day)` shown in **blue**.


## What is the average daily activity pattern?
```{r timeseriesMax, echo=TRUE}
g <- ggplot(data_interval, aes(x=interval, y=intervalCount))
g +
        geom_line() +
        xlim(0, 24) +
        xlab("Time of day, in hours") +
        ylab("Average Number of Steps in a 5-minute interval") +
        labs(title="Daily Activity Pattern")

max_interval <- data_interval[which.max(data_interval$intervalCount), ]
max_interval[1] <- tochar(max_interval[1])
```

The interval with the maximum number of steps on average is the five minute interval starting at `r max_interval[1]` AM with `r max_interval[2]` steps.


## Imputing missing values
```{r imputing, echo=TRUE}
NACount <- sum(is.na(data$steps))
fraction <- NACount / dim(data)[1]

# reprocessing the data, this time imputing the missing values
impData <- data
for(i in which(is.na(impData$steps))) {
        interval <- totime(impData$interval[i])
        impData$steps[i] <- data_interval$intervalCount[data_interval$interval==interval]
}
```

There are a total of `r NACount` missing values in the dataset, which represent `r fraction` of the total.

The data is imputed by replacing the missing value with the mean number of steps in that interval. The new data set is called *impData*.

Below is a histogram similar to what we saw before.
```{r hist2, echo=TRUE}
impData_day <- impData %>%
                group_by(date) %>%
                summarise(dailyCount=sum(steps, na.rm=TRUE))

mean.impDay <- mean(impData_day$dailyCount)
median.impDay <- median(impData_day$dailyCount)
impDayStats <- round(c(mean.impDay, median.impDay), 0)

g <- ggplot(impData_day, aes(x=dailyCount))
g +
        geom_histogram() +
        scale_x_log10() +
        geom_vline(xintercept=mean.impDay, size=2, color="red") +
        geom_vline(xintercept=median.impDay, size=2, color="blue") +
        xlab("Number of steps") +
        ylab("Frequency") +
        labs(title="Histogram of Steps Taken (Imputed Data)")
```

In the above plot, the mean of the number of steps taken is `r as.integer(impDayStats[1])` shown in **red** and the median of the number of steps taken is `r as.integer(impDayStats[2])` shown in **blue**.


## Are there differences in activity patterns between weekdays and weekends?
```{r, echo=TRUE, fig.height=10}
weekends <- c("Saturday", "Sunday")
isWeekday <- as.factor(!weekdays(impData$date) %in% weekends)
impData$isWeekday <- isWeekday

# subsetting data
impDataWeekday <- impData[impData$isWeekday==TRUE, ]
impDataWeekend <- impData[impData$isWeekday==FALSE, ]

# processing data
weekday_interval <- impDataWeekday %>%
                group_by(interval) %>%
                summarise(intervalCount=mean(steps, na.rm=TRUE))
weekday_interval$interval <- totime(weekday_interval$interval)

weekend_interval <- impDataWeekend %>%
                group_by(interval) %>%
                summarise(intervalCount=mean(steps, na.rm=TRUE))
weekend_interval$interval <- totime(weekend_interval$interval)

# plotting
par(mfrow=c(2, 1))
with(weekday_interval, plot(interval, intervalCount, type="l", xlab="",
                            ylab="On Weekdays"), lwd=3)

with(weekend_interval, plot(interval, intervalCount, type="l", xlab="Time of day, in hours",
                            ylab="On Weekends"), lwd=3)
```